Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-31 draft.tex
Added changes from Manuel

Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} play a major role for analyzing epidemiological data.
%
Graphic data representation is largely used to present results rather than gaining insight.

@@

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks labeling the voxels of an anatomical structure would be ideal since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
%
Information derived by landmarks, such as top and bottom point are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
%Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
%This gains even more momentum when analyzing multiple time steps.
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determining the type of tisse).

@@

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that enable a Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
%These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
Distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
We incorporate the idea of combining 3D shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann and colleagues identify local deformation changes by investigating shape-related difference on rodent mandibles \cite{Hermann2014}.
% TODO: Blöde formulierung
Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
%Covariance tensors indicate how a model is deformed when the user given a user specified deformation.
%The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%

@@

This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org/}}, which also validate the data (find missing data, clean up bad formatting, transform scales).

@@

A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables, the meaning and unit of measurement of their values.

@@

Morphometric features are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed \emph{IVA Loop} as seen in Figure~\ref{fig:WorkflowComparison}. Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types. The data are visualized in the range and domain space, which can then be brushed, yielding new groups, which can be investigated using further features. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The Sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{age} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
%
An hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection can also be derived from a shape-based clustering, which creates shape groups.
%
\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
%
The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
%
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
%
Subsequently the visualization can be either brushed or new features can be added to the analysis.
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
%
%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
%
%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
%
A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

@@

\paragraph{Pivot Tables.}
Pivot tables are frequently used to present the data in epidemiological publications.
%
Epidemiologists are used to process groups based on table representations.
%
Thus we decided to introduce an interactive pivot table.
%When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
%
%As seen in Figure~\ref{fig:visualization} (c) it is a good way to display how many subjects are in each group.
These tables clearly convey the subject count in each group (see Figure~\ref{fig:visualization} (c)).
%
However, they quickly get confusing and cluttered when they are divided into many subgroups.
%
We tackled this problem by making the order and number of displayed variables adaptable.
%
This also applies to the assignment of row or column variables.
%
Another way to avoid clutter is the user-driven selection of displayed variables.
%
To allow better comparison with respect to features, the values of each cell can also be displayed as percentage of the feature represented of either the row or the column.
% TODO IMPLEMENT
%The mean shape for each resulting sub-group is also displayed for each subject.

@@

\subsection{Implementation}
\begin{figure}[htb]
 \centering
 \label{fig:technologies}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/technologies}
 \caption{The front-end solution (left) uses state of the art web technologies such as \emph{HTML5}/\emph{CSS3}, \emph{WebGL} and \emph{SVG} to display the data.
 % 
 The \emph{NodeJS} based back-end (right) stores all image- and non-image data and transfers it to connected clients.
 %
 All computation-heavy operations, such as calculation of mean-shapes or -distances, as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
 %
 Client-server communication is accomplished via the \emph{Websocket} protocol.
 }
\end{figure}
%
To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies.
%
In addition to the obvious advantages of web technologies, the following aspects are crucial for our work:
\begin{itemize}
%	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
	\item The client-server structure allows for employing heavy computation on a server machine and transfer results to the client.
	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
	%
	High confidentiality standards of the data can be met by restricting access via a account system
	\item Recent developments in \emph{WebGL} applications running in browsers with near native performance push the development into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities. We use \emph{WebGL} for rendering shape information.
%	\item Web technologies are free.
\end{itemize}
These advantages do not come without drawbacks.
%
%Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
Due to missing ports of sophisticated specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{www.vtk.org}} or \texttt{R}\footnote{Open Source; \url{www.r-project.org}} for statistics, many standard methods need to be written from scratch for the web.
%Many methods, which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics, have build in, need to be written from scratch in order to fit in the context.
%

The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \url{nodejs.org}}, which is based on the Google V8 Javascript runtime environment.
%
Due to its event-driven non-blocking I/O model it is fast and does not freeze in case of heavy workload like mesh calculation.
\\\\
Non-image data for all subjects including the data dictionary is stored in a JSON file on the Server.
%
Image data are available as raw DICOM files as well as meshes representing segmentation masks of anatomical structures, which can be used to compare subjects.
%
The requested data is transmitted when a client connects.
%
The server processes calculation heavy statistical tasks such as calculation of \emph{Cram\'{e}rs' V} values for all feature combinations in order to keep the computation time on the client as low as possible.
%

The front-end is created using \texttt{Twitter Bootstrap}\footnote{Developed by Twitter, \url{getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%
Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
%
The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
%
\texttt{WebGL} rendering is carried out using the \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{threejs.org}} and allows GPU-accelerated data rendering.
%
Communication between client and server is enabled via the WebSockets protocol.
%
Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{www.mathworks.com}}, we hat to access them using the \texttt{NodeJS} server.
%
We accomplished this by converting it to a parameterized standalone console application, that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard-out and returns it in a proper format to the client.
%
All parameter-steered console applications can be incorporated in this context.
\\\\
The following section describes how our presented \emph{IVA} workflow works for an epidemiological use case.

\section{Application}
We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
%
It is one of the most common reasons for an adult to see a physician in the western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby and colleagues \cite{Harreby1996}, is largely focused on non-image information.
%
%If at all, only a few shape-related features are included in comparable studies, for example by Lang and colleagues \cite{Lang2011}.
%In comparable studies, only a few shape-related features are included, for example by Lang and colleagues \cite{Lang2011}.
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
% Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
% \begin{itemize}
% 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
% 	\item prognostic features for diagnosis and treatment of lumbar back pain
% 	\item determination of particularly effected risk groups
% \end{itemize}
%
Characterizing the healthy aging process of the spine is a long term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.

@@

\paragraph{Image Data.} \label{Image-Data}
The lumbar spine was detected in the image data using a hierarchical finite element method by Rak and colleagues \cite{Rak2013}.
%
This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L3-vertebra.
%
Two user-defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
%Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
%
It uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
%
The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
%
Due to incorrect initialization, strongly deformed spines, contrast-differences and artifacts, the model was not able to detect lumbar spines for all subjects.
%
We obtained and work with 983 tetrahedron models of the lumbar spine.
%
For clustering purposes, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).

\subsection{Shape Visualization and Clustering}
%The detection algorithm described in Section~\ref{Data Preprocessing} yields a model that determines voxel belonging to the lumbar spine.
%
The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
%
This allows for calculation of shape-distance and similarity.
%
This information is used to calculate mean-shapes as described in Section~\ref{Interaction- and Visualization Techniques}.

Shape distance is mapped onto color.
%
For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestation it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
\\\\
Shape-based clustering is carried out via Agglomerative Hierarchical clustering of the spine canal centerlines, (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
%
Since the "correct" number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
%Since it is not possible to determine the number of clusters in a given group, it is automatically computed using a \emph{knee/elbow} point which is described as tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
%
%For details, see \cite{Klemm2013VMV}.
%
The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.

>>>>

Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{http://ibm.com/software/analytics/spss/}} play a major role for analyzing epidemiological data.
%
Graphic data representation is largely used to present results rather than gaining insight.

@@

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks representing the voxels of an anatomical structure would be ideal since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
%
Information derived by landmarks, such as top and bottom point are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
%Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
%This gains even more momentum when analyzing multiple time steps.
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determining the type of tissue).

@@

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{http://tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{http://spotfire.tibco.com}} provide a rich user interface that enable a Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
%These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
Distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
We incorporate the idea of combining 3D shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann and colleagues identify local deformation changes by investigating shape-related difference on rodent mandibles \cite{Hermann2014}.
% TODO: Blöde formulierung
Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
%Covariance tensors indicate how a model is deformed when the user given a user specified deformation.
%The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%

@@

This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org}}, which also validate the data (find missing data, clean up bad formatting, transform scales).

@@

%A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables and the meaning and unit of measurement of their values
A data dictionary stores information about each manifestation of a feature.
Detailed description of data variables, its meaning as well as unit of measurement are stored as a lookup table.

@@

Morphometric features are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed \emph{IVA Loop} as seen in Figure~\ref{fig:WorkflowComparison}.
 Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
 The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features.
 % The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features. 
 Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The Sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{age} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
An hypothesis-driven analysis usually starts with the selection of features.
A feature selection from a shape-based clustering, which creates shape groups.
\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
Subsequently the visualization can be either brushed or new features can be added to the analysis.
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

@@

\paragraph{Pivot Tables.}
Pivot tables are frequently used to present the data in epidemiological publications.
%
Epidemiologists are used to process groups based on table representations.
%
Thus we decided to introduce an interactive pivot table.
%When opening a random epidemiological paper the reader will in almost all cases find some sort of pivot table to present the data.
%
%As seen in Figure~\ref{fig:visualization} (c) it is a good way to display how many subjects are in each group.
These tables clearly convey the subject count in each group (see Figure~\ref{fig:visualization} (c)).
%
However, they quickly get confusing and cluttered when they are divided into many subgroups.
%
We tackled this problem by making the order and number of displayed variables adaptable.
%
This also applies to the assignment of row or column variables.
%
Another way to avoid clutter is the user-driven selection of displayed variables.
%
To allow better comparison with respect to features, the values of each cell can also be displayed as percentage of the feature represented of either the row or column.
% TODO IMPLEMENT
%The mean shape for each resulting sub-group is also displayed for each subject.

@@

\subsection{Implementation}
\begin{figure}[htb]
 \centering
 \label{fig:technologies}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/technologies}
 \caption{The front-end solution (left) uses state of the art web technologies such as \emph{HTML5}/\emph{CSS3}, \emph{WebGL} and \emph{SVG} to display the data.
 % 
 The \emph{NodeJS} based back-end (right) stores all image- and non-image data and transfers it to connected clients.
 %
 All computation-heavy operations, such as calculation of mean shapes or -distances, as well as statistical processing is done by the server to keep hardware requirements of client systems low. 
 %
 Client-server communication is accomplished via the \emph{Websocket} protocol.
 }
\end{figure}
%
To provide a fast communication loop between method development and expert input, we decided to rely on modern web technologies.
%
In addition to the obvious advantages of web technologies, the following aspects are crucial for our work:
\begin{itemize}
%	\item No additional software needs to be installed, most people use a decent state-of-the-art web browser, even on mobile devices.
	\item The client-server structure allows for employing heavy computation on a capable server machine and transfer results to the client.
	\item Since image data for several thousand subjects requires hundreds of gigabytes disk space it can remain safely on the server and elements can be transferred on demand.
	%
	High confidentiality standards of the data can be met by restricting access via an account system
	\item Recent developments in \emph{WebGL} applications running in browsers with near native performance push the development further into to the web resulting in many open source libraries, which are well documented, rich in examples and driven by active communities. We use \emph{WebGL} for rendering shape information.
%	\item Web technologies are free.
\end{itemize}
These advantages do not come without drawbacks.
%
%Many methods which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics have build in, need to be written from scratch in order to fit in the context.
Due to missing ports of sophisticated and specialized libraries/languages, such as the \texttt{Visualization Toolkit} (\texttt{VTK})\footnote{Developed by Kitware Inc; \url{http://vtk.org}} or \texttt{R}\footnote{Open Source; \url{http://r-project.org}} for statistics, many standard methods need to be written from scratch for the web.
%Many methods, which specialized libraries, like the Visualization Toolkit (VTK) or R for statistics, have build in, need to be written from scratch in order to fit in the context.
%

The back-end is written using \texttt{NodeJS}\footnote{Developed by Joyent Inc, \url{http://nodejs.org}}, which is based on the Google V8 Javascript runtime environment.
%
Due to its event-driven non-blocking I/O model it is fast and does not freeze in case of heavy workload like mesh calculation.
\\\\
Non-image data for all subjects including the data dictionary is stored in a JSON file on the server.
%
Image data are available as raw DICOM files as well as meshes representing segmentation masks of anatomical structures, which can be used to compare subjects.
%
The requested data is transmitted when a client connects.
%
The server performs heavy statistical tasks such as calculation of \emph{Cram\'{e}rs' V} values for all feature combinations in order to keep the computation time on the client as low as possible.
%

The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \url{http://getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%
Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
%
The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
%
\texttt{WebGL} rendering is carried out by \texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} and allows GPU-accelerated data rendering.
%
Communication between client and server is enabled via the WebSockets protocol.
%
Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we hat to access them using the \texttt{NodeJS} server.
%
We accomplished this by converting it to a parameterized standalone console application, that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard-out and returns it in a proper format to the client.
%
All parameter-steered console applications can be incorporated in this context.
\\\\
%The following section describes how our presented \emph{IVA} workflow works for an epidemiological use case.
The following section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
\section{Application}
We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
%
It is one of the most common reasons for an adult to see a physician in the western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby and colleagues \cite{Harreby1996}, is largely focused on non-image information.
%
%If at all, only a few shape-related features are included in comparable studies, for example by Lang and colleagues \cite{Lang2011}.
%In comparable studies, only a few shape-related features are included, for example by Lang and colleagues \cite{Lang2011}.
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
% Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
% \begin{itemize}
% 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
% 	\item prognostic features for diagnosis and treatment of lumbar back pain
% 	\item determination of particularly effected risk groups
% \end{itemize}
%
Characterizing the healthy aging process of the spine is a long term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.

@@

\paragraph{Image Data.} \label{Image-Data}
The lumbar spine was detected in the image data using a hierarchical finite element method by Rak and colleagues \cite{Rak2013}.
%
This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L3-vertebra.
%
Two user-defined landmarks on the top and bottom of the L3-vertebra are used to obtain an initial height estimation of the model.
%Two more user- defined landmarks are the top and bottom of the L3-vertebra, which transforms size of the model accordingly to match the L3-vertebra .
%
It uses a weighted sum of T1- and T2-weighted MR images to detect the lumbar spine shape.
%
The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae \cite{Klemm2013VMV}.
%
Due to incorrect initialization, strongly deformed spines, contrast-differences and artifacts, the model was not able to detect lumbar spines for all subjects.
%
We obtained and worked with 983 tetrahedron models of the lumbar spine.
%
For clustering purposes, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature \cite{Klemm2013VMV}).

\subsection{Shape Visualization and Clustering}
%The detection algorithm described in Section~\ref{Data Preprocessing} yields a model that determines voxel belonging to the lumbar spine.
%
The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
%
This allows for calculation of shape-distance and similarity.
%
This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.

Shape distance is mapped onto color.
%
For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
\\\\
Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines, (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
%
Since the "correct" number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
%Since it is not possible to determine the number of clusters in a given group, it is automatically computed using a \emph{knee/elbow} point which is described as tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
%
%For details, see \cite{Klemm2013VMV}.
%
The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.

\subsection{Exploratory Analysis of the Lumbar Spine Data Set}
%
%We analyzed the given dataset expert-guided to assess its suitability for supporting both hypothesis-free analysis as well as generation.
Expert-guided analysis assessed the suitability of our approach for supporting both hypothesis-free analysis as well as hypothesis generation. 
%
The use of \emph{IVA} patterns as presented in Section~\ref{Image Centric Cohort Study Data in Interactive Visual Analysis Context} is denoted accordingly.
%
\subsubsection{Hypothesis-Free Analysis} \label{Hypothesis free Analysis}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/hypothesisfree}
 \caption{(a) Clustering result of all subjects. The bar chart height indicates the number of subjects in the cluster. The difference to the mean shape is color-coded where blue represents no difference and red a large difference. (b) A pivot table relates cluster results to gender and employment status. Cluster 8 is striking because of the high proportion of women, while Cluster 2, 3, 4 contain almost only men.}
 \label{fig:hypothesisfree}
\end{figure}
%
Analyzing the data set without a prior hypothesis requires a starting point, giving an overview over the data first \cite{Shneiderman1996}.
%
%Shape-based hypothesis-free exploration starts with a shape-grouping step achieved by shape-based clustering.
Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape-grouping step using shape-based clustering.
%
The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).
